---
title: "My puppy_Report"
execute: 
  warning: false
  error: false
  echo: false      # change to true to show the code
code-fold: false # change to true to fold the code chunks
---

# Motivation

In a world where dogs are considered not just pets but beloved family members, understanding what drives the popularity of certain breeds offers a unique lens into broader societal trends.
This project began with a simple question: Why do some dog breeds surge in popularity while others quietly fade from the spotlight?
As dog ownership continues to rise across the U.S.—particularly among millennials and Gen Z—shifting breed preferences reflect larger lifestyle changes, including increased urban living, smaller households, and a growing emphasis on emotional well-being.

Beyond curiosity, this project has real-world relevance.
For pet industry professionals, breeders, shelters, and prospective dog owners, identifying the traits linked to breed popularity can inform smarter decisions, from breeding strategies to adoption campaigns.
Culturally, it invites us to consider how our evolving relationships with animals mirror the ways our lives and values are changing.

# Research Question

This project explores shifting trends in dog breed popularity in the United States, aiming to uncover which breeds are gaining or losing favor over time and why. It investigates whether consistent behavioral or physical traits correlate with these popularity changes, providing insight into broader societal preferences. By applying machine learning techniques such as K-means clustering, the project seeks to group breeds based on shared characteristics and identify latent categories that appeal to different segments of the population. Ultimately, the goal is to reveal meaningful patterns in how Americans relate to dogs—and how that relationship is evolving over time.

# Background

To interpret the findings of this analysis, it is essential to understand several key concepts.

The American Kennel Club (AKC) publishes annual breed popularity rankings based on registration data, where a lower rank indicates higher popularity. Additionally, the AKC provides trait scores for each breed—such as Trainability, Energy Level, and Affectionate With Family—on a standardized 1 to 5 scale.

Popularity Shift is defined as the change in a breed’s ranking between 2013 and 2020. A positive value denotes increased popularity over time, while a negative value indicates a decline.

This project also employs K-means clustering, a machine learning technique used to group breeds based on trait similarities. This method enables the identification of latent categories that may not align with traditional breed groupings, offering deeper insight into evolving preferences in dog ownership.

# Data

-   **Data collection**

This project uses three main datasets: breed_traits.csv, which includes various characteristics of dog breeds such as friendliness, energy level, and grooming needs; trait_description.csv, which provides explanations for each trait; and breed_rank.csv, which contains AKC (American Kennel Club) popularity rankings for dog breeds from 2013 to 2020.

The popularity data covers an eight-year period from 2013 to 2020. The trait data does not include specific timestamps but represents general, widely accepted characteristics of each breed.

These datasets were collected for the TidyTuesday community to support practice in data visualization, analysis, and exploratory learning around dog breed trends and traits.

The breed rankings were sourced directly from official AKC reports. The breed traits and trait descriptions were scraped or compiled by contributors to the project.

The full dataset was curated by TidyTuesday organizers, primarily by data scientist Thomas Mock and other contributors from the R for Data Science community.

-   **Data Acquisition**

    -   Where / how did you get the data?
        -   The data was accessed via the tidytuesdayR R package or downloaded directly from GitHub.
    -   What is the source?
        -   TidyTuesday GitHub (2022-02-01)
        -   Original AKC rankings: AKC Website

-   **Data Understanding**

    -   How much data do you have?
        -   breed_traits.csv: 195 dog breeds × 23 traits
        -   breed_rank.csv: Rankings for 195 breeds across 8 years (2013–2020)
        -   trait_description.csv: 23 trait descriptions
    -   What types of measurements?
        -   Categorical/ordinal traits (e.g., “Very Low” to “Very High”)
        -   Yearly rankings (numeric)
        -   Text descriptions of traits
    -   Anything you needed to clean before getting started?
        -   Trait values were text-based and needed to be converted to ordered factors or numerical scales for analysis.
        -   Some breeds had minor inconsistencies in naming between datasets and required matching.
        
        
        
## America's Dog Preferences Are Changing: Who's Gaining — and Who's Dropping — in Popularity? {.unnumbered}

Between 2013 and 2020, America’s taste in dogs shifted notably, with certain breeds climbing the popularity ranks and others quietly fading from favor. A recent data analysis from the American Kennel Club’s breed rankings reveals telling trends in canine preferences, hinting at evolving lifestyle needs, aesthetic tastes, and family dynamics.


```{r}
# Get the Data

# read in the data manually

breed_traits <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-02-01/breed_traits.csv')
trait_description <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-02-01/trait_description.csv')
breed_rank_all <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-02-01/breed_rank.csv')

library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)

head(breed_traits)

breed_rank_all_clean <- breed_rank_all %>%
  filter(if_all(c(`2013 Rank`, `2014 Rank`, `2015 Rank`, `2016 Rank`,
                  `2017 Rank`, `2018 Rank`, `2019 Rank`, `2020 Rank`), ~ !is.na(.)))



```


### Part 1: overall trend {.unnumbered}
```{r}
breed_rank_trend <- breed_rank_all_clean %>%
  mutate(Rank_Change = as.numeric(`2013 Rank`) - as.numeric(`2020 Rank`)) %>%
  arrange(desc(Rank_Change)) 
```

```{r}
breed_rank_trend <- breed_rank_all_clean %>%
  mutate(Rank_Change = as.numeric(`2013 Rank`) - as.numeric(`2020 Rank`))

top_gainers <- breed_rank_trend %>% 
  slice_max(Rank_Change, n = 5)

top_losers <- breed_rank_trend %>% 
  slice_min(Rank_Change, n = 5)

gainers_long <- top_gainers %>%
  pivot_longer(cols = starts_with("20"), names_to = "Year", values_to = "Rank") %>%
  mutate(Year = str_remove(Year, " Rank") %>% as.integer())

ggplot(gainers_long, aes(x = Year, y = as.numeric(Rank), color = Breed)) +
  geom_line(size = 1.2) +
  scale_y_reverse() +
  labs(title = "Top 5 Rising Dog Breeds (2013–2020)",
       y = "Popularity Rank (Lower is More Popular)",
       x = "Year") +
  theme_minimal() +
  theme(legend.position = "bottom")

losers_long <- top_losers %>%
  pivot_longer(cols = starts_with("20"), names_to = "Year", values_to = "Rank") %>%
  mutate(Year = str_remove(Year, " Rank") %>% as.integer())

ggplot(losers_long, aes(x = Year, y = as.numeric(Rank), color = Breed)) +
  geom_line(size = 1.2) +
  scale_y_reverse() +
  labs(title = "Top 5 Falling Dog Breeds (2013–2020)",
       y = "Popularity Rank (Lower is More Popular)",
       x = "Year") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

A line chart tracking the top five rising dog breeds shows a clear upward trajectory in popularity between 2013 and 2020. Breeds such as the Beaucerons, Belgian Malinois, and  Cane Corso experienced significant gains, with some jumping more than 20 spots in national ranking.

These breeds tend to be medium to large in size, loyal, quiet, and have high scores in affection and trainability—characteristics increasingly favored by urban middle class and suburban pet owners.

#### More Part 1 {.unnumbered}
```{r} 
rank_traits <- breed_rank_all_clean %>%
  left_join(breed_traits, by = "Breed")

rank_traits <- rank_traits %>%
  mutate(Rank_Change = as.numeric(`2013 Rank`) - as.numeric(`2020 Rank`),
         Trend = case_when(
           Rank_Change > 0 ~ "Rising",
           Rank_Change < 0 ~ "Falling",
           TRUE ~ "Stable"
         ))

trait_score_cols <- c("Affectionate With Family", "Good With Young Children",
                      "Good With Other Dogs", "Shedding Level", "Coat Grooming Frequency",
                      "Drooling Level", "Openness To Strangers", "Playfulness Level",
                      "Watchdog/Protective Nature", "Adaptability Level",
                      "Trainability Level", "Energy Level", "Barking Level", 
                      "Mental Stimulation Needs")

long_trait <- rank_traits %>%
  filter(Trend %in% c("Rising", "Falling")) %>%
  select(Breed, Trend, all_of(trait_score_cols)) %>%
  pivot_longer(cols = -c(Breed, Trend), names_to = "Trait", values_to = "Score") %>%
  mutate(Score = as.numeric(Score))

trait_summary <- long_trait %>%
  group_by(Trend, Trait) %>%
  summarize(Average_Score = mean(Score, na.rm = TRUE), .groups = "drop")

ggplot(trait_summary, aes(x = Trait, y = Average_Score, fill = Trend)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Trait Comparison: Rising vs. Falling Dog Breeds",
       y = "Average Trait Score", x = "Trait") +
  theme_minimal() +
  theme(legend.position = "top")
```

Conversely, breeds like the Treeing walkers, Chinooks, and Spaniels have seen notable declines in popularity. The line chart for the top five falling breeds reveals a steady downward drift.
Common traits among these breeds include higher grooming needs, increased drooling, and lower adaptability scores, suggesting that maintenance-heavy dogs are becoming less appealing in modern households.



### The Traits of the Top (and Bottom) Dogs
Continue with previous visualizations, we know what type of dogs are popular and have made a summary of them, but what are their specific characteristics?
In the chart titled “Trait Comparison: Top 10 vs. Bottom 10 Dog Breeds,” we compared the behavioral and physical traits of the most popular breeds with those at the bottom of the popularity list.

```{r}
rank_long <- breed_rank_all %>%
  pivot_longer(cols = ends_with("Rank"),
               names_to = "year",
               values_to = "rank") %>%
  mutate(
    year = str_remove(year, " Rank"),
    year = as.integer(year)
  )


latest_year <- max(rank_long$year, na.rm = TRUE)

latest_ranks <- rank_long %>%
  filter(year == latest_year) %>%
  select(Breed, rank)


breed_data <- breed_traits %>%
  left_join(latest_ranks, by = "Breed")

top_bottom <- breed_data %>%
  filter(!is.na(rank)) %>%
  mutate(rank_group = case_when(
    rank <= 10 ~ "Top 10",
    rank >= max(rank, na.rm = TRUE) - 9 ~ "Bottom 10",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(rank_group))


numeric_traits <- top_bottom %>%
  select(where(is.numeric), -rank) %>%
  names()

trait_long <- top_bottom %>%
  select(Breed, rank_group, all_of(numeric_traits)) %>%
  pivot_longer(cols = -c(Breed, rank_group), names_to = "trait", values_to = "score") %>%
  group_by(rank_group, trait) %>%
  summarise(mean_score = mean(score, na.rm = TRUE), .groups = "drop")


```


```{r}


ggplot(trait_long, aes(x = reorder(trait, mean_score), y = mean_score, fill = rank_group)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Trait Comparison: Top 10 vs. Bottom 10 Dog Breeds",
       x = "Trait",
       y = "Average Score",
       fill = "Breed Popularity") +
  theme_minimal(base_size = 13)

```

A few standout trends emerged:
Top 10 breeds (like the Beaucerons and Belgian) scored much higher on:
Affectionate With Family
Good With Children
Trainability
Playfulness
Bottom 10 breeds were more likely to have:
Higher Shedding Levels
Greater Grooming Needs
Lower Scores for Adaptability and Energy
This suggests a clear preference for dogs that are easy to care for, eager to please, and great with families. Breeds that are more high-maintenance or aloof — even if charming in their own right — may be falling out of step with modern lifestyles.

### Dividing dogs' characteristics to suitable owners {.unnumbered}

#### K-means/elbow method/clusters {.unnumbered}
K-Means Clustering: This is a machine learning method used to group similar items (in this case, dog breeds) based on multiple numeric traits. It helps uncover hidden groupings or "types" of dogs that might not align with traditional breed groupings.
Popularity Shift (Rank Change): Defined as the difference in a breed’s ranking between 2013 and 2020. A positive value indicates rising popularity, and a negative value shows decline.

```{r}

library(cluster)  
library(factoextra) 


rank_long <- breed_rank_all %>%
  pivot_longer(cols = ends_with("Rank"), 
               names_to = "year", 
               values_to = "rank") %>%
  mutate(year = str_remove(year, " Rank"),
         year = as.integer(year))


latest_ranks <- rank_long %>%
  filter(year == max(year, na.rm = TRUE)) %>%
  select(Breed, rank)


breed_data <- breed_traits %>%
  left_join(latest_ranks, by = "Breed") %>%
  filter(!is.na(rank))


trait_data <- breed_data %>%
  select(where(is.numeric)) %>%
  select(-rank) %>%  
  scale()  


fviz_nbclust(trait_data, kmeans, method = "wss") + 
  labs(title = "Elbow Method to Choose Number of Clusters")


```

```{r}

set.seed(123)  
k <- 4
km <- kmeans(trait_data, centers = k, nstart = 25)


breed_clustered <- breed_data %>%
  mutate(cluster = factor(km$cluster))

# trait-based clustering
fviz_cluster(km, data = trait_data,
             geom = "point",
             main = "Dog Breed Clusters Based on Traits",
             labelsize = 8,
             palette = "Set2")


breed_clustered %>%
  group_by(cluster) %>%
  summarise(avg_rank = mean(rank, na.rm = TRUE),
            n_breeds = n()) %>%
  ggplot(aes(x = cluster, y = avg_rank, fill = cluster)) +
  geom_col() +
  labs(title = "Average Popularity Rank by Trait Cluster",
       x = "Cluster",
       y = "Average Rank (lower = more popular)") +
  theme_minimal(base_size = 13)


```

Each cluster = a behavioral or personality profile.
You can describe them as:

The active adventurers The chill lap dogs The obedient family dogs The aloof independents

 Cluster 1: High trainability, affectionate, moderate energy → "Family-Friendly, Obedient Breeds"
 Cluster 2: Super high energy, low trainability → "High-Drive Working Dogs"

️ Cluster 3: Low energy, low shedding → "Laid-Back Companions"

 Cluster 4: High grooming needs, moderate traits → "Show Dogs or Fancy Breeds"

Analyzing the clusters 2-D dimensions visualization and the average popularity rank by traits together we can know, cluster 1 and 3 stand out. 
Cluster 1 (Family Dogs) had by far the lowest average rank, making them the most popular group.
Cluster 2 (Adventurers) and Cluster 4 (Show Dogs) lagged behind, possibly due to their high energy or grooming requirements.
Cluster 3 (Lap Dogs) occupied the middle ground, offering a good compromise for many owners.

#### Traits summary per cluster {.unnumbered}

```{r}


cluster_traits_summary <- breed_clustered %>%
  mutate(cluster = factor(cluster)) %>%
  group_by(cluster) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))


print(cluster_traits_summary)



```

### Conclusion {.unnumbered}

In the evolving world of pet ownership, America's favorite dogs reflect more than just cuteness or tradition — they mirror our values, lifestyles, and changing daily realities. Through breed popularity data and trait analysis, a clear narrative emerges: the most beloved dog breeds today are those that are loyal, affectionate, adaptable, and easy to live with.

Machine learning deepened that story, revealing four distinct canine "personality clusters" — from the reliable, trainable family dogs to the independent show dogs and high-energy workers. The most popular breeds overwhelmingly came from clusters that emphasized emotional connection, low maintenance, and versatility, aligning with trends toward urban living, work-from-home routines, and active family dynamics.

Ultimately, this project underscores a simple truth: as our lives evolve, so do our choices in companionship. Today’s top dogs aren’t just pets — they’re partners, confidants, and reflections of who we are.
